Equalizer\\Programming and User Guide

### 2012-01-23 Developer/ProgrammingGuide/paper.tex
Partial update for 1.2 Programming Guide

\lstset{language=C++}
\section*{Equalizer 1.0 Programming and User Guide}
\today
\subsection*{Contributors}

@@

\textcopyright 2007-2011 \htmladdnormallink{Eyescale Software GmbH}
{http://www.eyescale.ch}. All rights reserved. No permission is
granted to copy, distribute, or create derivative works from the
contents of this electronic documentation in any manner, in whole or in
part, without the prior written permission of Eyescale Software GmbH.

@@

Equalizer is the most advanced middleware for scalable 3D visualization,
providing the broadest set of parallel rendering features available in
an open source library to any OpenGL application. Many commercial and
open source applications in a variety of different markets rely on
Equalizer for flexibility and scalability.

@@

\fig{fExecutionFlow} illustrates the basic principle of any parallel
rendering application. The typical OpenGL application, for example GLUT,
has an event loop which redraws the scene, updates data based on
received events, and eventually redraws a new frame.

A parallel rendering application uses the same basic execution model and
extends it by separating the rendering code from the main event
loop. The rendering code is then executed in parallel on different
resources, depending on the configuration chosen at runtime.

@@

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and compiling the source code. After installing Equalizer, please take a
look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of Equalizer and the
\textsf{eqPly} example.

Equalizer uses CMake to generate platform-specific build files. Compiling
Equalizer is as simple as running \textsf{make} on Linux or Mac OS X, running
\textsf{make xcode} and building the generated XCode project on Mac OS X, or
creating a Equalizer Visual Studio solution using \textsf{CMake} on Windows.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

Collage, the underlying network library, uses a peer-to-peer communication
structure. Equalizer uses Collage to provide optimal performance scalable
rendering, e.g., by setting up direct communication between two nodes when
needed for image compositing or software swap barriers.

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
only supports one application per server, but it will provide
concurrent and efficient multi-application support in the future.

@@

The application connects to an Equalizer server and receives a
configuration.  Furthermore, the application also provides its render
client, which will be controlled by the server. The application reacts
on events, updates its data and controls the rendering.

@@

The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer. It executes the rendering tasks received from the server by
calling the appropriate task methods (see \sref{sTaskMethods}) in the
correct thread and context. The application either implements the task
methods with application-specific code or uses the default methods
provided by Equalizer.

@@

@@

Real-time visualization is an inherently parallel
problem. Unfortunately, different applications have different rendering
algorithms, which require different scalable rendering modes to address
the bottlenecks correctly. Equalizer supports all important algorithms as
listed below, and will continue to add new ones over time to meet
application requirements.

@@

2D decomposes the rendering in screen-space, that is, each contributing
rendering unit processes a tile of the final view. The recomposition
simply assembles the tiles side-by-side on the destination view. This
mode is also known as sort-first or SFR.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{ A 2D Compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if
possible. In most cases, a \textsf{loadBalancer} should be used to
automatically adjust the tiling each frame, based on the current
rendering load.

@@

Volume rendering applications use an ordered alpha-based blending to
composite the result image. The depth buffer information is used to
composite the individual images correctly for polygonal data.

@@

Unfortunately, the database recomposition has linear increasing IO
requirements for the pixel transfer. Parallel recomposition algorithms,
such as direct-send address this problem by keeping the per-node IO
constant (see \fig{fDirectSend}).

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}\small A Database Compound}}
\end{wrapfigure}
The application has to partition the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

@@

Due to the frame consistency between the eye views, this modes scales
very well. The IO requirements for pixel transfer are small and
constant.

The number of rendering resources used by stereo compounds is limited by
the number of eye passes, typically two.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
\end{wrapfigure}
Stereo compounds are used by all applications when rendering in stereo,
and is often combined with other modes.

@@

DPlex compounds assign full, alternating frames to individual rendering
units. The resulting images are copied to the destination channel, and
Equalizer load-balancing is used to ensure a steady framerate on the
destination window. This mode is also known as time-multiplex or AFR.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{ A DPlex Compound}}
\end{wrapfigure}

@@

DPlex compounds in Equalizer are configured using the \textsf{period}
and \textsf{phase} parameter, limiting each child to render a subset of
the frames. Each child compound uses an output frame, which is connected
to an input frame on the destination channel. The destination channel
uses a DPlex load-balancer to smoothen the framerate.

@@

As 2D compounds, pixels compounds have low, constant IO requirements for
the pixel transfers during recomposition. They are naturally load-balanced for
fill-limited operations, while they do not decompose the geometry processing at
all.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A Pixel Compound}}
\end{wrapfigure}
OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at runtime to adjust the rendering accordingly.

Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources. Pixel compounds are ideal for ray-tracing,
which is highly fill-limited and needs the full database for rendering
anyway. Volume rendering applications are also well suited for this
mode, and should choose it over 2D compounds.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
\end{wrapfigure}
Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

@@

Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

As for the DB compound, the subpixel recomposition has linear increasing 
IO requirements for the pixel transfer with the difference that only 
color information has to be transmitted.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
\end{wrapfigure}
Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

Please note that the ZeroConf module is optional and requires a ZeroConf
implementation when Equalizer is compiled.

@@

\subsubsection{Usage}

@@

\begin{itemize}
\item The default value is \textsf{local} which uses all local GPUs queried
  using the cgl, glx or wgl GPU-SD module.
\item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
  GPU-SD to query all GPUs in the subnet for the given session. The default
  session of the GPU-SD daemon is \textsf{default}. The ZeroConf hostname is
  used to connect the nodes.
\item \textsf{--eq-config filename.eqc} loads a static configuration from the
  given ascii file. The following sections describe how to write configuration
  files.
\end{itemize}

@@

A configuration consists of the declaration of the rendering resources
and the usage description for these resources. It can optionally contain
the description of the physical layout of the projection system and
logical layouts on the projection canvases.

@@

The resource usage is configured using a compound tree, which is a
hierarchical representation of the rendering decomposition and
recomposition across the resources. It is explained in \sref{sCompounds}.

Physical layouts of display systems are configured using canvases with
segments, which represent 2D rendering areas composed of multiplex
displays or projectors. Logical layouts are applied to canvases and
define views on a canvas.

@@

\fig{fConfig} shows an example configuration for a four-side
CAVE, running on two machines (nodes) using three graphics
cards (pipes) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

@@

For each machine in your cluster, create one \textsf{node}. Create one
\textsf{appNode} if your application process should also render, which
is often the case. List all nodes, even if you are not planning to use
them at first. Equalizer will only instantiate and access used nodes,
that is, nodes which are referenced by an active compound.

@@

@@

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. They
should only be used if using a different, synchronized rendering thread
is not an option.

@@

\fig{fFrusta} illustrates the wall and projection frustum parameters.

Canvases provide a convenient way to configure projection surfaces. A
canvas uses layouts, which describe logical views. The usage of canvases
and layouts is optional. Typically, each desktop window uses one canvas.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}
A canvas represents one logical projection surface, e.g., a PowerWall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

@@

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A Canvas using four Segments}}
\end{wrapfigure}

@@

@@

An observer represents an actor looking at multiple views. It has a head
matrix, defining its position and orientation within the world, and an
eye separation. Typically, a configuration has one
observer. Configurations with multiple observers are used if multiple,
head-tracked users are in the same configuration session, e.g., a
non-tracked control host with two tracked head-mounted displays.

@@

Compound trees are used to describe how the rendering resources are
combined to produce the desired output, and how they are aggregated to
increase the performance.

It is advised to study and understand the basic configuration files
shipped with the Equalizer configuration, before attempting to write
compound configurations. The command line program \textsf{configTool},
shipped with the Equalizer distribution, creates some standard
configurations automatically.

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level
  compound.
  \begin{description}
  \item[Standalone channels] Set the channel by using the channel name
    from the resource section. Define a frustum as explained in
    \sref{sCanvas}.
  \item[View/Segment channels] Set the channel by using the canvas, segment,
    layout and view name. The compound frustum will be calculated automatically
    based on the segment or view frustum. Note that one segment may created
    multiple view/segment channels, one for each view intersection of each
    layout used on the canvas. Only the compounds belonging to the active layout
    of a canvas are active at runtime.
    \begin{description}
    \item [Software swap lock] When using software swap synchronization,
      swap-lock the destination compounds using a swap barrier. All
      windows with a swap barrier of the same name synchronize their
      swapbuffers.
    \item [Hardware swap lock] When using hardware swap synchronization,
      use swap barriers for the destination compounds, setting
      \textsf{NV\_group} and \textsf{NV\_barrier} appropriately. The
      swap barrier name is ignored in this case. All windows of the same
      group on a single node synchronize their swap buffer. All groups
      of the same barrier synchronize their swap buffer across
      nodes. Please note that the driver typically limits the number of
      groups and barriers to one, and that multiple windows per swap
      group are not supported by all drivers. Please refer to your
      OpenGL driver documentation for details.
    \end{description}
  \end{description}
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task of that
    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

\subsubsection{Compound Channels}
Each compound has a channel, which is used by the compound to execute
the rendering tasks. One channel might be used by multiple
compounds. Unused channels, windows, pipes and nodes are not
instantiated during config initialization. The rendering tasks for the
channels are computed by the server and send to the appropriate render
client nodes at the beginning of each frame.

@@

Compounds have a frustum description to define the physical layout of
the display environment. The frustum specification is described in
\sref{sCanvas}.

The frustum description is inherited by the children, therefore the
frustum is typically defined on the topmost compound.

@@

\subsubsection{Tasks}
Compounds execute a number of tasks: clear, draw, assemble and
readback. By default, a leaf compound executes all tasks and a non-leaf
compound assemble and readback. A non-leaf compound never executes the
draw task.

@@

\subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
Compounds may have a software or hardware swap barrier to synchronize the
buffer swap of multiple channels. A software swap barrier is configured
by giving it a name. Windows with a swap barrier of the same name
synchronize with each other before executing the swap buffer
task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.

Note that a software swap barrier is not accurate enough for
edge-blended projection system or immersive installations. Such
installations typically use a hardware synchronization, e.g., nVidia
G-Sync cards. It is however sufficient for display walls made out of
LCD's and for some non-immersive projection systems.

Swap barrier names are only valid within the compound tree, that is, a
compound from one compound tree cannot be synchronized with a compound
from another compound tree.

A hardware swap barrier is configured by setting the \textsf{NV\_group} and
\textsf{NV\_barrier} parameters. These parameters follow the
\textsf{NV\_swap\_group} extension, which synchronizes all windows bound to the
same group on a single machine, and all groups bound to the same barrier across
systems.

>>>>

\lstset{language=C++}
\section*{Equalizer 1.2 Programming and User Guide}
\today
\subsection*{Contributors}

@@

\textcopyright 2007-2012 \htmladdnormallink{Eyescale Software GmbH}
{http://www.eyescale.ch}. All rights reserved. No permission is
granted to copy, distribute, or create derivative works from the
contents of this electronic documentation in any manner, in whole or in
part, without the prior written permission of Eyescale Software GmbH.

@@

Equalizer is the most advanced middleware for scalable 3D visualization,
providing the broadest set of parallel rendering features available in
an open source library to any visualization application. Many commercial and
open source applications in a variety of different markets rely on
Equalizer for flexibility and scalability.

@@

\fig{fExecutionFlow} illustrates the basic principle of any parallel rendering
application. The typical OpenGL application, for example using GLUT, has an
event loop which redraws the scene, updates application data based on received
events, and eventually renders a new frame.

A parallel rendering application uses the same basic execution model, extending
it by separating the rendering code from the main event loop. The rendering code
is then executed in parallel on different resources, depending on the
configuration chosen at runtime.

@@

Equalizer can be installed by downloading a binary installer or source
code\footnote{\link{http://www.equalizergraphics.com/downloads.html}}. After
installing Equalizer, please take a look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of Equalizer and the \textsf{eqPly}
example.

Equalizer uses CMake to generate platform-specific build files. Compiling
Equalizer from source is as simple as running \textsf{make} on Linux or Mac OS
X, running \textsf{make xcode} and building the generated XCode project on Mac
OS X, or creating a Equalizer Visual Studio solution using \textsf{CMake} on
Windows.

@@

The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
Collage is a cross-platform C++ library for building heterogenous, distributed
applications. Collage provides an abstraction of different network connections,
peer-to-peer messaging, discovery and synchronization as well as
high-performance, object-oriented, versioned data distribution. Collage is
designed for low-overhead multi-threaded execution which allows applications to
easily exploit multi-core architectures. Equalizer uses Collage as the cluster
backend, e.g., by setting up direct communication between two nodes when needed
for image compositing or software swap barriers.

\fig{fProcesses} depicts the relationship between the server, application,
render client and adminstrative processes, which are explained below.

\subsubsection{Server}
The Equalizer server is responsible for managing one visualization session on a
shared memory system or graphics cluster. Based on its configuration and
controlling input from the application, it computes the active resources,
updates the configuration and generates tasks for all processes. Furthermore it
controls and launches the application's rendering client processes. The
Equalizer server is the entity in charge of the configuration, and all other
processes receive their configuration from the server.

@@

The application connects to an Equalizer server and receives a configuration.
Furthermore, the application also provides its render client, which will be
controlled by the server. The application and render client may use the same
executable. The application has a main loop, which reacts on events, updates its
data and controls the rendering.

@@

The render client implements the rendering part of an application. Its execution
is passive, it has no main loop and is completely driven by Equalizer server. It
executes the rendering tasks received from the server by the calling the
appropriate task methods (see \sref{sTaskMethods}) in the correct thread and
context. The application either implements the task methods with
application-specific code or uses the default methods provided by Equalizer.

@@

\subsubsection{Administration Programs}

Equalizer 1.0 introduced the admin library, which can be used to modify a
running Equalizer server. The admin library is still in development, but already
allows limited modifications such as adding new windows and changing
layouts. The admin library may be used to create standalone administration tools
or from within the application code. In any case, it has an indepedent view of
the server's configuration. Documentation for admin library is not yet part of
this Programming and User Guide.

@@

Scalable rendering is a subset of parallel rendering, where more multiple
resources are used to update a view.

Real-time visualization is an inherently parallel problem. Different
applications have different rendering algorithms, which require different
scalable rendering modes to address the bottlenecks correctly. Equalizer
supports all important algorithms as listed below, and will continue to add new
ones over time to meet application requirements.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{A 2D Compound}}
\end{wrapfigure}
2D decomposes the rendering in screen-space, that is, each contributing
rendering unit processes a tile of the final view. The recomposition
simply assembles the tiles side-by-side on the destination view. This
mode is also known as sort-first or SFR.

The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if
possible. In most cases, a \textsf{load\_equalizer} should be used to
automatically adjust the tiling each frame, based on the current
rendering load.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}\small A Database Compound}}
\end{wrapfigure}
Volume rendering applications use an ordered alpha-based blending to
composite the result image. The depth buffer information is used to
composite the individual images correctly for polygonal data.

@@

Unfortunately, the database recomposition has linear increasing IO
requirements for the pixel transfer. Parallel compositing algorithms,
such as direct-send, address this problem by keeping the per-node IO
constant (see \fig{fDirectSend}).

The application has to partition the database so that the rendering units render
only part of the database. Some OpenGL features do not work correctly
(anti-aliasing) or need special attention (transparency, shadows).

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
\end{wrapfigure}
Due to the frame consistency between the eye views, this modes scales very
well. The IO requirements for pixel transfer are small and constant. The number
of rendering resources used by stereo compounds is limited by the number of eye
passes, typically two.

Stereo compounds are used by all applications when rendering in stereo,
and is often combined with other modes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2em}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{ A DPlex Compound}}
\end{wrapfigure}
DPlex compounds assign full, alternating frames to individual rendering
units. The resulting images are copied to the destination channel, and
Equalizer load-balancing is used to ensure a steady framerate on the
destination window. This mode is also known as time-multiplex or AFR.

@@

DPlex compounds in Equalizer are configured using the \textsf{period}
and \textsf{phase} parameter, limiting each child to render a subset of
the frames. Each child compound uses an output frame, which is connected
to an input frame on the destination channel. The destination channel
uses a \textsf{framerate\_equalizer} to smoothen the framerate.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A Pixel Compound}}
\end{wrapfigure}
As 2D compounds, pixel compounds have low, constant IO requirements for the
pixel transfers during recomposition. They are naturally load-balanced for
fill-limited operations, but do not scale geometry processing at all.

OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at runtime to adjust the rendering accordingly.

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
\end{wrapfigure}
Pixel compounds work well for purely fill-limited applications. Techniques like
view frustum culling do not reduce the rendered data, since each resource has
approximately the same frustum. Pixel compounds are ideal for ray-tracing, which
is highly fill-limited and needs the full database for rendering anyway. Volume
rendering applications are also well suited for this mode, and should choose it
over 2D compounds.

Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
\end{wrapfigure}
Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

As for the DB compound, the subpixel recomposition has linear increasing 
IO requirements for the pixel transfer, with the difference that only 
color information has to be transmitted.

Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[~index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

Please note that both GPU-SD and the GPU-SD ZeroConf module are optional and
requires GPU-SD and a ZeroConf implementation when Equalizer is compiled.

@@

\subsubsection{\label{sGPUSDUsage}Usage}

@@

\begin{itemize}
\item The default value is \textsf{local} which uses all local GPUs queried
  using the cgl, glx or wgl GPU-SD module.
\item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
  GPU-SD to query all GPUs in the subnet for the given session. The default
  session of the GPU-SD daemon is \textsf{default}. The hostname provided by
  ZeroConf is used to connect the nodes.
\item \textsf{--eq-config filename.eqc} loads a static configuration from the
  given ascii file. The following sections describe how to write configuration
  files.
\end{itemize}

@@

A configuration consists of the declaration of the rendering resources, the
description of the physical layout of the projection system, logical layouts on
the projection canvases and an optional decomposition description using the
aforementioned resources.

@@

Physical layouts of display systems are configured using canvases with
segments, which represent 2D rendering areas composed of multiple
displays or projectors. Logical layouts are applied to canvases and
define views on a canvas.

Scalable resource usage is configured using a compound tree, which is a
hierarchical representation of the rendering decomposition and
recomposition across the resources.

@@

\fig{fConfig} shows an example configuration for a four-side
CAVE, running on two machines (nodes) using three graphics
cards (pipes) with one window each to render to the four output channels
connected to the projectors for each of the walls.

@@

For each machine in your cluster, create one \textsf{node}. Create one
\textsf{appNode} for your application process. List all nodes, even if you are
not planning to use them at first. Equalizer will only instantiate and access
used nodes, that is, nodes which are referenced by an active compound.

@@

For cluster configurations with multiple nodes, configure at least one
connection for the server. All render clients connect back to the server, for
which this connection is needed.

@@

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU), and
uses one independent operating system thread for rendering. Non-threaded pipes
are supported for integrating with thread-unsafe libraries, but have various
performance caveats. They should only be used if using a different, synchronized
rendering thread is not an option.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}
\fig{fFrusta} illustrates the wall and projection frustum parameters.

A canvas represents one physical projection surface, e.g., a PowerWall, a
curved screen or an immersive installation. One configuration might
drive multiple canvases, for example an immersive installation and an
operator station.

@@

A canvas may have a swap barrier, which is used as the default swap barrier by
all its segments to synchronize the video output.

Canvases provide a convenient way to configure projection surfaces. A canvas
uses layouts, which describe logical views. Typically, each desktop window uses
one canvas, one segment, one layout and one view.

@@

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A Canvas using four Segments}}
\end{wrapfigure}
To synchronize the video output, configure either a canvas swap barrier or a
swap barrier on each segment to be synchronized.

When using software swap synchronization, swap-lock all segments using a swap
barrier. All windows with a swap barrier of the same name synchronize their
swapbuffers. Software swap synchronization uses a distributed barrier, and works
on all hardware.

When using hardware swap synchronization, use swap barriers for all segment to
be synchronized, setting \textsf{NV\_group} and \textsf{NV\_barrier}
appropriately. The swap barrier name is ignored in this case. All windows of the
same group on a single node synchronize their swap buffer. All groups of the
same barrier synchronize their swap buffer across nodes. Please note that the
driver typically limits the number of groups and barriers to one, and that
multiple windows per swap group are not supported by all drivers. Hardware swap
barriers require support from the OpenGL driver, and has been tested on nVidia
Quadro GPUs with the G-Sync option. Please refer to your OpenGL driver
documentation for details.

@@

An observer represents an actor looking at multiple views. It has a head matrix,
defining its position and orientation within the world, an eye separation and
focus distance parameters. Typically, a configuration has one
observer. Configurations with multiple observers are used if multiple,
head-tracked users are in the same configuration session, e.g., a non-tracked
control host with two tracked head-mounted displays.

@@

Compound trees are used to describe how multiple rendering resources are
combined to produce the desired output, especially how multiple GPUs are
aggregated to increase the performance.

It is advised to study and understand the basic configuration files shipped with
the Equalizer configuration, before attempting to write compound
configurations. The auto-configuration code and command line program
\textsf{configTool}, shipped with the Equalizer distribution, creates some
standard configurations automatically. These are typically used as templates for
custom configuration files.

@@

\subsubsection{Writing Compounds}
The following steps are typically taken when writing compound
configurations:
\begin{description}
\item [Root compound] Define an empty top-level compound when
  synchronizing multiple destination views. Multiple destination views
  are used for multi-display systems, e.g., a PowerWall or CAVE. All
  windows used for one display surface should be swap-locked (see below)
  to provide a seamless image. A single destination view is typically
  used for providing scalability to a single workstation window.
\item [Destination compound(s)] Define one compound for each destination
  channel, either as a child of the empty group, or as a top-level compound. Set
  the destination channel by using the canvas, segment, layout and view name or
  index. The compound frustum will be calculated automatically based on the
  segment or view frustum. Note that one segment may created multiple
  view/segment channels, one for each view intersection of each layout used on
  the canvas. Only the compounds belonging to the active layout of a canvas are
  active at runtime.
\item[Scalability] If desired, define scalability for each of your
  destination compounds. Add one compound using a source channel for
  each contributor to the rendering. The destination channel may also be
  used as a source. 
  \begin{description}
  \item[Decomposition] On each child compound, limit the rendering task of that
    child by setting the \textsf{viewport}, \textsf{range}, \textsf{period} and
    \textsf{phase}, \textsf{pixel}, \textsf{subpixel}, \textsf{eye} or
    \textsf{zoom} as desired.
  \item[Runtime Adjustments] A \textsf{load\_equalizer} may be used on
    the destination compounds to set the \textsf{viewport} or
    \textsf{range} of all children each frame, based on the current
    load. A \textsf{view\_equalizer} may be used on the root compound to
    assign resources to all destination compounds, which have to use
    \textsf{load\_equalizer}s. A \textsf{framerate\_equalizer} should be
    used to smoothen the framerate of DPlex compounds. A
    \textsf{DFR\_equalizer} may be used to set the zoom of a compound to
    achieve a constant framerate. One compound may have multiple
    equalizers, e.g., a \textsf{load\_equalizer} and a
    \textsf{DFR\_equalizer} for a 2D compound with a constant framerate.
  \item[Recomposition] For each source compound, define an
    \textsf{output\_frame} to read back the result. Use this output frame
    as an \textsf{input\_frame} on the destination compound. The frames
    are connected with each other by their name, which has to be unique
    within the root compound tree. For parallel compositing, describe
    your algorithm by defining multiple input and output frames across
    all source compounds.
  \end{description}
\end{description}

\subsubsection{Compound Channels}
Each compound has a channel, which is used by the compound to execute the
rendering tasks. One channel might be used by multiple compounds. Compounds are
only active if their corresponding destination channel is active, that is, if
the parent layout of the view which created the destination channel is active on
at least one canvas.

Unused channels, windows, pipes and nodes are not instantiated during
initialization.  Switching an active layout may cause rendering resources to be
stopped and started. The rendering tasks for the channels are computed by the
server and send to the appropriate render client nodes at the beginning of each
frame.

@@

Compounds have a frustum description to define the physical layout of the
display environment. The frustum specification is described in
\sref{sCanvas}. The frustum description is inherited by the children, therefore
the frustum is defined on the topmost compound, typically by the corresponding
segment.

@@

\subsubsection{Tasks}
Compounds execute a number of tasks: clear, draw, assemble and readback. By
default, a leaf compound executes all tasks and a non-leaf compound assemble and
readback. A non-leaf compound will never execute the draw task.

@@

\subsubsection{\label{sSwapBarrier}Synchronization - Swap Barriers}
Compounds may have a software or hardware swap barrier to synchronize the buffer
swap of multiple channels. Swap barriers are defined by the segment, which will
set the swap barrier on all its destination channel compounds.

A software swap barrier is configured by giving it a name. Windows with a swap
barrier of the same name synchronize with each other before executing the swap
buffer task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.  Swap barrier names are only
valid within the compound tree, that is, a compound from one compound tree
cannot be synchronized with a compound from another compound tree.

A hardware swap barrier uses a hardware component to synchronize the buffer swap
of multiple windows. It guarantees that the swap happens at the same vertical
retrace of all corresponding video outputs.  It is configured by setting the
\textsf{NV\_group} and \textsf{NV\_barrier} parameters. These parameters follow
the \textsf{NV\_swap\_group} extension, which synchronizes all windows bound to
the same group on a single machine, and all groups bound to the same barrier
across systems.

Display synchronization uses different algorithms. Framelock synchronizes each
vertical retrace of multiple graphic outputs using a hardware component. This is
configured in the driver, independently of the application. Genlock synchronizes
each horizontal and vertical retrace of multiple graphic outputs, but is not
used anymore for 3D graphics. Swap lock synchronizes the front and back buffer
swap of multiple windows, either using a software solution based on network
communication or a hardware solution based on a physical cable. It is
independent of, but often used in conjunction with framelock.

Framelock is used to synchronize the vertical retrace in multi-display active
stereo installations, e.g., for edge-blended projection systems or immersive
installations. It is combined with a software or hardware swap barrier. Software
barriers in this case can not guarantee that the buffer swap of all outputs
always happens with the same retrace. The time window for an output to miss the
retrace for the buffer swap is however very small, and the output will simply
swap on the next retrace, typically in 16 milliseconds.
